Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-09-14 paper/paper.tex
add maximal benchmark

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{bb}\toprule
            \hline
            \textbf{element type}     & \textbf{purpose} \\
            {real number}     & {counting all independent sets} \\
            {tropical number} (\Eq{eq:countingtropical})    & {finding the independence number} \\
            {tropical number with counting} (\Eq{eq:countingtropical})     & {finding the independence number and the number of MISs} \\
            {tropical number with configurations} (\Eq{eq:singleconfig})     & {finding the independence number and one MIS} \\
            {tropical number with sets} (\Eq{eq:countingtropicalset})    & {finding the independence number and enumeration of all MISs} \\
            {polynomial} (\Eq{eq:polynomial})     & {computing the independence polynomial exactly} \\
            {truncated polynomial} (\Eq{eq:max2poly})     & {counting MISs and independent sets of size $\alpha(G)-1$} \\
            {complex number}     & {fitting the independence polynomial with fast Fourier transform} \\
            {finite-field algebra (\Eq{eq:finitefield})} & {fitting the independence polynomial exactly using number theory} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and their purposes in calculating various independent set properties.}\label{tbl:generictypes}
\end{minipage}
\red{we should include maximal independence polynomial and counting, enumeration in this table as well.}
\end{table}

@@

\section{Maximum independent sets and its counting problem}
\subsection{Tropical algebra for finding the independence number and counting MISs}
In the previous section, we focused on computing the independence polynomial for a graph $G$ of given independence number $\alpha(G)$, but we did not show how to compute this number.
The method we use to compute this quantity is based on the following observations. Let $x=\infty$, the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where the lower-order terms vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with $a(k) = (a_{k}, k)$, where the second element stores the largest exponent $k$ and the first element stores the corresponding coefficient $a_{k}$.
From this, we can define a new algebra as
%\begin{equation}
%\begin{split}
%    a(k) \oplus a(j) &= \begin{cases}
%        \left(a_k + a_j, \max(k,j) \right), & k = j \\
%        \left(a_j, \max(k,j) \right), & k < j \\
%        \left(a_k, \max(k,j) \right), & k > j
%    \end{cases}, \\
%    a(k) \odot a(j) &= (a_k a_j, k+j) \\
%    \mymathbb{0} &= (0, -\infty)\\
%    \mymathbb{1} &= (1, 0). \label{eq:countingtropical}
%\end{split}
%\end{equation}
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
In the program, we only need to store the exponent $x$ and the corresponding coefficient $a_x$ initialized to $1$. 
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}.:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}
\end{equation}
This algebra is the same as the one used in Liu et al.~\cite{Liu2021} to calculate and count spin glass ground states.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm tropical} = \left(\begin{matrix}
        \mymathbb{1} \\
        (1,1)
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm tropical} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

@@

>>>>

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
            \textbf{element type}     & \textbf{purpose} \\
            {$\mathbbm{R}$}     & {real number for counting all independent sets} \\
            {$\mathbbm{C}$}     & {complex number for fitting the independence polynomial with fast Fourier transform} \\
            {$\text{GF}(p)$} (\Eq{eq:finitefield}) & {number on finite field algebra for fitting the independence polynomial exactly using number theory} \\
            {T} (\Eq{eq:tropical})    & {tropical number for finding the independence number} \\
            {P1} (\Eq{eq:countingtropical})     & {polynomial truncated to largest order, i.e. the tropical number with counting field for finding the independence number and the number of MISs} \\
            {P2} (\Eq{eq:max2poly})     & {counting MISs and independent sets of size $\alpha(G)-1$} \\
            {PN} (\Eq{eq:polynomial})     & {computing the independence polynomial exactly} \\
            {P1+S1}     & {replacing the counting field in P1 to configuration sampler defined in $\Eq{eq:singleconfig}$ for finding the independence number and one MIS} \\
            {P1+SN} (\Eq{eq:countingtropicalset})    & {replacing the counting field in P1 to configuration sampler defined in $\Eq{eq:set}$ for finding the independence number and enumeration of all MISs} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and their purposes in calculating various independent set properties.}\label{tbl:generictypes}
\end{minipage}
\red{we should include maximal independence polynomial and counting, enumeration in this table as well.}
\end{table}

@@

\section{Maximum independent sets and its counting problem}
\subsection{Tropical algebra for finding the independence number and counting MISs}
In the previous section, we focused on computing the independence polynomial for a graph $G$ of given independence number $\alpha(G)$, but we did not show how to compute this number.
The method we use to compute this quantity is based on the following observations. Let $x=\infty$, the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where the lower-order terms vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with $a(k) = (a_{k}, k)$, where the second element stores the largest exponent $k$ and the first element stores the corresponding coefficient $a_{k}$.
From this, we can define a new algebra as
%\begin{equation}
%\begin{split}
%    a(k) \oplus a(j) &= \begin{cases}
%        \left(a_k + a_j, \max(k,j) \right), & k = j \\
%        \left(a_j, \max(k,j) \right), & k < j \\
%        \left(a_k, \max(k,j) \right), & k > j
%    \end{cases}, \\
%    a(k) \odot a(j) &= (a_k a_j, k+j) \\
%    \mymathbb{0} &= (0, -\infty)\\
%    \mymathbb{1} &= (1, 0). \label{eq:countingtropical}
%\end{split}
%\end{equation}
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
In the program, we only need to store the exponent $x$ and the corresponding coefficient $a_x$ initialized to $1$. 
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}.:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
This algebra is the same as the one used in Liu et al.~\cite{Liu2021} to calculate and count spin glass ground states.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm tropical} = \left(\begin{matrix}
        \mymathbb{1} \\
        (1,1)
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm tropical} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig2.pdf}
    \caption{Benchmark results for computing different properties of maximal independent sets on a random three regular graph with different tensor element types.
    (a) treewidth versus the number of vertices for the benchmarked graphs. 
    (b) The computing time for calculating the number of independent sets and enumerate all MISs.
    }
    \label{fig:benchmark-maximal}
\end{figure}

We also benchmarked our method with the Bron-Kerbosch method to enumerate and count maximal independent sets in \Fig{fig:benchmark-maximal}.
We can see the tree width of this tensor network is nearly doubled comparing with the one for computing independent sets, hence only small graphs can be benchmarked.
The time for the tensor netwrok approach and Bron-Kerbosch approach to enumerate all maximal independent sets are comparable,
while the counting can be much faster.
